A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-04 podc_paper/algorithm.tex
working on proof (mainly indexdequeue)

An operation can terminate only after a block containing it has been appended to the root's blocks array.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.

@@

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $v$, we do not yet know where its
superblock will eventually be installed in $B$'s parent.
So, we leave $B.\fld{super}$ blank.  After $B$ is installed 
in $v.\fld{blocks}[\var{i}]$, processes cooperate to fill in $B.\fld{super}$ 
when they attempt to advance $v.\fld{head}$ from $i$ to $i+1$,
using a value they read from the \fld{head} field of $v$'s parent.
As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
shall prove that it is close.

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, has been }
    \State \linecomment{propagated to root and contains at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
	    \EndIf
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\opemph{GetEnqueue}(\var{v,b,i}) returns the argument of the
$i$th enqueue in the $b$th block of \typ{Node} $v$. 
The range of subblocks of a block is determined using the \fld{end\sub{left}} and
\fld{end\sub{right}} fields of the block and its previous block. Then,
the exact subblock is found using binary search on the \fld{sum\sub{enq}}
field (line \ref{getChild}). 
\here{explain a little more?}

>>>>

We shall show that an operation terminates only after a block containing it has been appended to the root's blocks array.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.

@@

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $v$, we do not yet know where its
superblock will eventually be installed in $B$'s parent.
So, we leave $B.\fld{super}$ blank.  After $B$ is installed 
in $v.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
when they attempt to advance $v.\fld{head}$ from $h$ to $h+1$,
using a value they read from the \fld{head} field of $v$'s parent.
As mentioned above, this might not be the exact index of $B$'s superblock, but we
shall prove that it is close.

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, has been }
    \State \linecomment{propagated to root and contains at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}
	    \EndIf
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
$i$th enqueue in the $b$th block of \typ{Node} $v$. 
The range of subblocks of a block is determined using the \fld{end\sub{left}} and
\fld{end\sub{right}} fields of the block and its previous block. Then,
the exact subblock is found using binary search on the \fld{sum\sub{enq}}
field (line~\ref{getChild}). 
\here{explain a little more?}

### 2023-01-04 podc_paper/correctness.tex
working on proof (mainly indexdequeue)

The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.

\begin{lemma}
  If $B=v.\var{blocks}[b]$ is a direct subblock of $v.\fld{parent.blocks}[s]$ then $B.\fld{super}\leq s$.
\end{lemma}
-------

\begin{proof}
Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$s$]} into $n\nf{.parent}$. By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}. By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}. The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}. From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq s$, since \nf{$n$.parent.head} is still equal to $s$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}. The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
\end{proof}

Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
\end{definition}

Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.

\begin{observation}\label{lem::createBlockHead}
If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
\end{observation}

\begin{lemma}\label{lem::secondRefreshSuper}
Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}. Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
\end{lemma}

\begin{proof}
Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.

Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$. From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.

$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}. Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}. We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}. Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}. 
\begin{align*}
\left.
\begin{array}{r}
 _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
\end{array}
\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
\end{align*}

So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.

Therefore $R_{p2}$ reads $n\nf{.head}=b$. $R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$. So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$. This is in contradiction with our hypothesis.

Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
\end{proof}

\begin{corollary} \label{superRelation}
If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
\end{corollary}
\begin{proof}
% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}. To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}. The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$. So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
\end{proof}

We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
\here{mention somewhere why precondition of IndexDequeue is true; also check that block of parent indexed by sup in that routine is non-null}

\begin{lemma}[\nf{IndexDequeue} correctness]
 If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}. After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}. By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right. If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}). Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}). Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis. 
\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}

  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child. The index of the superblock is shown with $sb$. }\label{fig::blockSumDeqLeft}
\end{figure}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}

  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child. The index of the superblock is shown with $sb$. }\label{fig::blockSumDeqRight}
\end{figure}

@@

\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}. By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.

>>>>

Next, we prove the \fld{super} field of a block is within one of the true index of the block's superblock.
\here{An alternative would be modify the code
so that the correct index of the superblock is computed before storing it in $B.super$.
In Advance, we could check if $h>v.parent.blocks[h_p].end_{left/right}$ and if so write $h_p+1$ instead of $h_p$ in $v.blocks[h].super$.  Then, in IndexDequeue there would be no need to correct the value read from the super field.  If we do this, the following lemma would be modified to prove that the value is perfectly accurate.}

\begin{lemma}\label{superRelation}
Let $B=v.\var{blocks}[b]$.
  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{lemma}
\begin{proof}
We first show $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($v.\fld{parent}$) that installs the superblock of $B$ 
in $v.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $v.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
By \Cref{lem::headPosition}, $B.\fld{super} \neq \nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $v.\fld{parent}.\head$ before $r$.
When $r$ occurs, $v.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by \Cref{lem::headPosition}, $v.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored $B.\fld{super}$ was read from $v.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume for the remainder that $s>2$.
By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(v.\fld{parent})$ that installed the block in $v.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $v.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $v$;
otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
$v.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($v, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $v.\fld{blocks}[h]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $v.\fld{blocks}[h]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By \Cref{nonDecreasingHead}, $v.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $v.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $v.\fld{parent}.\head$.
\end{proof}

%The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $
%\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block 
%into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.

We prove \op{IndexDequeue}'s correctness using \Cref{superRelation} on each step of the \op{IndexDequeue}.
\here{mention somewhere why precondition of IndexDequeue is true; also check that block of parent indexed by sup in that routine is non-null}

\begin{lemma}
If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
 then \op{IndexDequeue}($v, b, i$) returns $\langle b',i'\rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
\end{lemma}
\begin{proof}
We prove the claim by induction on the depth of node $v$. The base case where $v$ is the root is trivial (see Line \ref{indexBaseCase}).
Assuming the claim holds for $v$'s parent, we prove it for $v$.
Let $B=v.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($v, b, i$) first computes the index $sup$ of $B'$ in $v.\fld{parent}$.
By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $v.\fld{parent.blocks}[B.\fld{super}]$.

Next, the position of the required dequeue in $D(B')$ is computed in 
lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
We first add the number of dequeues in the subblocks of $B'$ in $v$ that preced $B$ on line \ref{computeISuperStart}.
If $v$ is the right child of its parent, then all of the subblocks of $B'$ from $v$'s left sibling
also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.

Finally, \op{IndexDequeue} is called recursively on $v$'s parent.
Since $B$ has been propagated to the root, so has its superblock $B'$.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required dequeue in the root.\end{proof}

----------

@@
